---
title: "Basic GWAS Plots"
vignette: >
  %\VignetteIndexEntry{Basic GWAS Plots}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

gwasplot provides fast Manhattan and QQ plots for GWAS summary statistics.
This vignette walks through the core plotting workflow using simulated data —
no real GWAS files or DuckDB connections required.

## Simulate data

The helper below generates a data frame with standard columns `CHROM`, `POS`,
and `PVALUE`, distributing variants proportionally across chromosomes and
optionally injecting genome-wide significant signals.

```{r simulate}
library(gwasplot)

# Approximate chromosome sizes (Mb, hg38)
chr_mb <- c(249, 242, 198, 190, 181, 171, 159, 145, 138, 134,
            135, 133, 115, 107, 102,  90,  83,  80,  59,  63, 47, 51)

sim_gwas <- function(n_total, hits = NULL,
                     chroms = paste0("chr", 1:22)) {
  sizes <- chr_mb[seq_along(chroms)]
  dfs <- Map(function(ch, mb) {
    n <- max(1L, round(n_total * mb / sum(sizes)))
    data.frame(
      CHROM  = ch,
      POS    = sort(sample.int(mb * 1e6L, n)),
      PVALUE = runif(n)
    )
  }, chroms, sizes)
  df <- do.call(rbind, dfs)
  for (h in hits) {
    idx <- df$CHROM == h$chrom & abs(df$POS - h$pos) < h$window
    df$PVALUE[idx] <- runif(sum(idx), h$pmin, h$pmax)
  }
  df
}

set.seed(42)
gwas_df <- sim_gwas(
  n_total = 500000,
  hits = list(
    list(chrom = "chr2",  pos = 135e6, window = 5e5, pmin = 1e-15, pmax = 1e-8),
    list(chrom = "chr5",  pos =  50e6, window = 5e5, pmin = 1e-12, pmax = 1e-8),
    list(chrom = "chr11", pos =  70e6, window = 5e5, pmin = 1e-20, pmax = 1e-8),
    list(chrom = "chr17", pos =  30e6, window = 5e5, pmin = 1e-10, pmax = 1e-8)
  )
)

head(gwas_df)
```

## Genomic inflation factor

`lambda_gc()` computes the genomic inflation factor from the PVALUE column.
A value close to 1 indicates well-controlled test statistics.

```{r lambda}
lambda_gc(gwas_df)
```

## QQ plot

`qqplot()` accepts a numeric vector of p-values and returns a ggplot object,
which renders inline in the document.

```{r qqplot, fig.width=5, fig.height=5}
qqplot(gwas_df$PVALUE)
```

## Manhattan plot

`manhattan()` saves the plot to a file via `ggsave`. Pass `width`, `height`,
and `dpi` through `...` to control output dimensions.

```{r manhattan, fig.alt="Manhattan plot of simulated GWAS data"}
outfile <- knitr::fig_path(".png")
dir.create(dirname(outfile), showWarnings = FALSE, recursive = TRUE)
manhattan(gwas_df, output = outfile, width = 10, height = 4, dpi = 150, base_size = 14)
knitr::include_graphics(outfile)
```

The four injected signals on chromosomes 2, 5, 11, and 17 are clearly visible
above the genome-wide significance threshold (−log₁₀ p ≈ 7.3).
