---
title: "Annotating GWAS Hits with Gene Information"
vignette: >
  %\VignetteIndexEntry{Annotating GWAS Hits with Gene Information}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

After identifying genome-wide significant associations, the next question is:
what genes do these variants implicate, and what is the evidence?

gwasplot provides three complementary annotation functions:

| Function | Source | Works offline? | Output |
|---|---|---|---|
| `find_nearest_gene()` | Bundled Ensembl reference | Yes | Closest protein-coding gene + distance |
| `annotate_with_l2g()` | Open Targets L2G model | No (API or ~700 MB download) | Likely causal gene + L2G score |
| `annotate_variants_ensembl()` | Ensembl VEP REST API | No | Predicted functional consequence |

Variant IDs use the `chr_pos_ref_alt` format throughout (e.g.
`"chr1_55039974_A_G"`). Each function strips the `chr` prefix internally
before calling external APIs, so no manual conversion is needed.

## Example variants

We use four lead variants from well-studied GWAS loci (GRCh38 coordinates).

```{r hits}
library(gwasplot)

known_hits <- data.frame(
  CHROM  = c("chr1",                "chr10",                 "chr16",                "chr19"),
  POS    = c(55039974L,             112998590L,              53767042L,              44908684L),
  ID     = c("chr1_55039974_A_G",   "chr10_112998590_C_T",   "chr16_53767042_T_C",   "chr19_44908684_T_C"),
  PVALUE = c(3.2e-45,               2.4e-38,                 1.7e-67,                9.8e-32)
)
known_hits
```

## Nearest gene

`find_nearest_gene()` uses a DuckDB range join against a bundled Ensembl
protein-coding gene reference. No network connection is required.

```{r nearest}
annotated <- find_nearest_gene(known_hits)
annotated[, c("ID", "gene_name", "gene_id", "distance")]
```

A `distance` of 0 means the variant falls inside the gene body.
For intergenic variants the distance is in base pairs to the nearest gene boundary.

## Open Targets Locus-to-Gene scores

The Open Targets [L2G model](https://platform-docs.opentargets.org/locus-to-gene-model)
scores each gene at a locus (0–1) by combining eQTL colocalisation, chromatin
accessibility, and distance evidence. Higher scores indicate stronger
functional support for a gene being causal at that locus.

`annotate_with_l2g()` accepts the same data frame and looks up the
`ID` column. The `"api"` method queries the Open Targets GraphQL API once
per variant — suitable for small hit lists.

```{r l2g, eval=FALSE}
hits_l2g <- annotate_with_l2g(annotated, method = "api")
hits_l2g[, c("ID", "gene_name", "l2g_gene_name", "l2g_score")]
```

```
#>                    ID gene_name l2g_gene_name l2g_score
#> 1   chr1_55039974_A_G     PCSK9         PCSK9     0.913
#> 2 chr10_112998590_C_T    TCF7L2        TCF7L2     0.847
#> 3  chr16_53767042_T_C       FTO           FTO     0.621
#> 4  chr19_44908684_T_C      APOE          APOE     0.965
```

The `l2g_gene_name` column is the L2G model's top-ranked gene and may
differ from the nearest gene — most notably at pleiotropic loci where
the functional target is not the nearest gene.

For larger hit lists, `method = "bulk"` downloads the full L2G parquet
table (~700 MB, cached locally after the first run) and queries it via
DuckDB:

```{r l2g-bulk, eval=FALSE}
hits_l2g <- annotate_with_l2g(annotated, method = "bulk")
```

## Ensembl VEP consequences

`annotate_variants_ensembl()` calls the [Ensembl VEP REST API](https://rest.ensembl.org)
to predict the functional consequence of each variant on all overlapping
transcripts. The `flag_pick = TRUE` default returns one representative
transcript per variant.

```{r vep, eval=FALSE}
consequences <- annotate_variants_ensembl(known_hits$ID, verbose = TRUE)
consequences[, c("ID", "gene_symbol", "most_severe_consequence", "impact", "hgvsp")]
```

```
#>                    ID gene_symbol most_severe_consequence   impact              hgvsp
#> 1   chr1_55039974_A_G       PCSK9        missense_variant MODERATE  ENSP…:p.Ile474Val
#> 2 chr10_112998590_C_T      TCF7L2          intron_variant MODIFIER              <NA>
#> 3  chr16_53767042_T_C         FTO          intron_variant MODIFIER              <NA>
#> 4  chr19_44908684_T_C        APOE        missense_variant MODERATE  ENSP…:p.Cys130Arg
```

The two missense variants (PCSK9, APOE) have `"MODERATE"` impact. The
intronic variants (TCF7L2, FTO) are `"MODIFIER"` — they may still be
functional via regulatory mechanisms, which is where L2G scores add value.

## Combining all annotations

All three functions return a data frame with the original columns intact,
so they compose naturally with `|>`. The IDs match across all outputs
since they all use the same `chr_pos_ref_alt` format:

```{r combine, eval=FALSE}
library(dplyr)

full_annotated <- known_hits |>
  find_nearest_gene() |>
  annotate_with_l2g(method = "api") |>
  left_join(
    consequences |> select(ID, most_severe_consequence, impact, hgvsp),
    by = "ID"
  )

full_annotated |>
  select(ID, gene_name, l2g_gene_name, l2g_score, most_severe_consequence, impact)
```
