---
title: "Identifying Independent Loci"
vignette: >
  %\VignetteIndexEntry{Identifying Independent Loci}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

After association testing it is common to identify a small set of independent
signals from a larger pool of genome-wide significant variants.
`identify_loci()` uses a greedy, distance-based algorithm: it picks the most
significant variant as a locus lead, assigns all variants within a specified
window to that locus, then repeats on the remaining unassigned variants.

## Simulate data with signal clusters

The simulation places three tight clusters of significant variants on
chromosomes 1, 3, and 4.

```{r simulate}
library(gwasplot)

# Approximate chromosome sizes (Mb, hg38)
chr_mb <- c(249, 242, 198, 190, 181, 171, 159, 145, 138, 134,
            135, 133, 115, 107, 102,  90,  83,  80,  59,  63, 47, 51)

sim_gwas <- function(n_total, hits = NULL,
                     chroms = paste0("chr", 1:22)) {
  sizes <- chr_mb[seq_along(chroms)]
  dfs <- Map(function(ch, mb) {
    n <- max(1L, round(n_total * mb / sum(sizes)))
    data.frame(
      CHROM  = ch,
      POS    = sort(sample.int(mb * 1e6L, n)),
      PVALUE = runif(n)
    )
  }, chroms, sizes)
  df <- do.call(rbind, dfs)
  for (h in hits) {
    idx <- df$CHROM == h$chrom & abs(df$POS - h$pos) < h$window
    df$PVALUE[idx] <- runif(sum(idx), h$pmin, h$pmax)
  }
  df
}

set.seed(123)
gwas_df <- sim_gwas(
  n_total = 100000,
  chroms  = paste0("chr", 1:5),
  hits    = list(
    list(chrom = "chr1", pos = 100e6, window = 2e6, pmin = 1e-20, pmax = 1e-8),
    list(chrom = "chr3", pos =  60e6, window = 2e6, pmin = 1e-15, pmax = 1e-8),
    list(chrom = "chr4", pos =  80e6, window = 2e6, pmin = 1e-12, pmax = 1e-8)
  )
)

nrow(gwas_df)
```

## Filter to genome-wide significant hits

`select_top_hits()` filters rows to those below a p-value threshold
(default 5×10⁻⁸).

```{r top-hits}
top_hits <- select_top_hits(gwas_df, threshold = 5e-8)
nrow(top_hits)
```

## Identify independent loci

`identify_loci()` adds two columns to the filtered data:

- `locus_id` — integer locus index (1, 2, 3, …)
- `is_lead` — `TRUE` for the most significant variant in each locus

```{r loci}
loci <- identify_loci(top_hits, window_kb = 500)
head(loci)
```

## Lead variant table

Filtering to lead variants gives one row per independent signal.

```{r leads}
leads <- loci |>
  dplyr::filter(is_lead) |>
  dplyr::select(locus_id, CHROM, POS, PVALUE)

leads
```

The three signals on chromosomes 1, 3, and 4 are recovered as expected.

## Adding gene annotations

When working with a DuckDB-backed `GWASFormatter` object,
`find_nearest_gene()` can annotate each variant with its closest gene,
adding `gene_name`, `gene_id`, and `distance` columns.
See `?find_nearest_gene` for details.
